---
title: "Manhattan vs Sentinel (Sept 2018 Portfolio)"
author: "Rahul Birmiwal"
output:
  html_document: default
  pdf_document: default
---


## DEPENDENCIES

  - ggplot
  - dplyr 
  - treemapify
  - knitr
  - readxl 
  - gridExtra



```{r setup}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)
library(gridExtra)
library(treemapify)
library(knitr)
library(readxl)

```


### Table of Contents

  1. 'Broad' Competitive Analysis 
      - Portfolio Distributions 
  2. 'Granular' Analysis 
      - Treemap Visualizations
      - Tables of Total Asset Allocation Weights By Company 

The purpose of this report is to do broad-picture qualitative portfolio analysis between Manhattan Life and its top-competition firms. Now in particular, this R file seeks to do a one-on-one comparison between Sentinel Security Life Insurance and Manhattan Life. One question of interest is why Sentinel is able to pay a liability rate of $4% on one of its product; while Manhattan only 3.1% -- i.e. what covariates in Sentinel's porfolio differ from MLIC such that they can pay at a higher rate? 


### Functions

  - port_reader: Used to read in the portfolio .xls files that were retrieved via Bloomberg PORT export 
  - summarize_by_sector: helper function that summarizes data used for the histogram plots 
  - heirarchy_fill: function used to 'populate' (fill down) the parent-child structure of the excel file 
  - process_data: function that takes in the full dataframe from heirarchy_fill(), filters by finding the underlying issuers 
   whose net portoflio weight lies in the top 10%, and also filters by the important market sectors used for the treemap 
  - plot_treemap: function that generates the treemap visualizations 

```{r}
port_reader <- function(fname) { 

  df = readxl::read_excel(fname, col_names = T , skip = 9) #starting row at row 9 
  #print("#####################")
  #print(df %>% head(3))
  cname <- df$X__1[[1]][1] #read in the company name 
  df <- df[,-1] #remove blank at top 
  
  #Rename key columns 
  colnames(df)[1] <- 'Sector'
  colnames(df)[2] <- 'Subsector'
  colnames(df)[3] <- 'Security'
  colnames(df)[4] <- 'X..Wgt'
  df$Sector <- as.character(df$Sector)

  #Cast to appropriate types and fix issue with number formatting 
  df$Subsector <- (as.character(df$Subsector))
  df$Security <- (as.character(df$Security))
  df$X..Wgt <- as.numeric(gsub(",","",df$X..Wgt))
  df$Maturity <- stringr::str_sub(df$Security,-2,-1)
  df$Maturity <- as.numeric(df$Maturity)
  return(list(cname, df)) #return List of <comapny name, its dataframe> 
} 

#Function does a group_by to generate statistical totals based on overarching sectors
#Needed for the bar chart overlays 
summarize_by_sector <- function(df) { 
  df <- df %>% group_by(Sector) %>% summarise(Weight=sum(X..Wgt), MktVal = sum(`Market Value`))
  return(df)
}




```



```{r}
heirarchy_fill <- function(df) {
  foundFirst <- FALSE 
  for (i in 1:nrow(df)){
         if (is.na(df$Sector[i]) | df$Sector[i] == '') {
              if (foundFirst == FALSE) { 
                foundFirst <-TRUE
                next 
              } else { 
                df$Sector[i]<-df$Sector[i-1] 
              }
         }
  }



  foundFirst <- FALSE 
  for (i in 1:nrow(df)){
         if (is.na(df$Subsector[i]) | df$Subsector[i] == '') {
              if (foundFirst == FALSE) { 
                foundFirst <-TRUE
                next 
              } else { 
                df$Subsector[i]<-df$Subsector[i-1] 
              }
         }
  }
  
  df <- df[!(is.na(df$Security) | df$Security==""), ]
  
  return(df)
}

```


```{r}
imp_sectors = c("Asset Backed Securities", 
"Banks", 
"Commericial MBS", 
"Communications", 
"Consumer, Cylical", 
"Consumer, Non-cyclical", 
"Energy", 
"Financial", 
"Government", 
"Industrial", 
"Mortgage Securities", 
"REITS", 
"Whole Loan Collateral CMO")



processData <- function(df, only_imp) { 
  fixed <- heirarchy_fill(df)

  if (only_imp == TRUE) { 
    fixed <- fixed[fixed$Sector %in% imp_sectors, ]  
  }
  fixed$RootEquity <- sapply(fixed$Security,function(x) strsplit(x, " ")[[1]][1])
  importantData <- fixed %>% group_by(Sector, Subsector, RootEquity ) %>% 
  summarise(equityWeight = sum(X..Wgt), avgMaturity = mean(Maturity) - 18) %>% filter(equityWeight >= max(0.1, quantile(equityWeight, 0.90) ))
  return(importantData)
}
```


```{r}
plot_treemap <- function(d, cname) { 
  ggplot(d, aes(area=equityWeight, fill=Sector, subgroup = Sector,  subgroup2 = Subsector, label=RootEquity)) +   geom_treemap() + 
  geom_treemap_subgroup_border() +  geom_treemap_subgroup2_border() + geom_treemap_subgroup2_text(place = "topright", grow = F, alpha = 0.8, colour = "black", fontface = "italic", size = 12) + 
  geom_treemap_text(colour = "white", place = "centre", reflow = T, size = 10, alpha=0.9, fontface='bold') + ggtitle(paste('Granular Breakdown For',cname )) + theme(plot.title = element_text(size=7)) 

}

```



```{r LOAD PSRT DATA}
psrt_data_list <- list() 
cnames <- list() 
files <- list.files(path=getwd(), pattern="*.xls", full.names=T, recursive=FALSE)
#files <- files[unlist(lapply(files, function(x) (grepl(firm_of_interest, x)==TRUE  | grepl('manh', x)==TRUE) ))]


# Dataframe that consists of all the 'summary' dataframes stacked ontop of each other 
master_df <- data.frame() 


for (i in 1:length(files)) { 
  res <- port_reader(files[i]) #read in the List<companyName, data> 
  cname <- (res[[1]]) #get the company name 
  print(cname)
  cnames[[i]] <- cname #store the company name 
  full_df <- (res[[2]])  #store dataframe in temp variable 

  summary_df <- summarize_by_sector(full_df) #do summary analysis for the plot 
  psrt_data_list[[i]] <- full_df #store the fll dataframe in a list 
  
  yield <- 4
  #Hardcoard a yield column into the dataframe based on appropriate company 
  if (grepl("ATLANTIC", cname) == TRUE) { 
    yield <- 4.00  
  }
  if (grepl("COLORADO", cname) == TRUE) { 
    yield <- 3.80  
  }
  if (grepl("FIDELITY", cname) == TRUE) { 
    yield <- 3.80 
  }
  if (grepl("LIBERTY", cname) == TRUE) { 
    yield <- 3.70 
  }
  if (grepl("MANHATTAN", cname) == TRUE) { 
    yield <- 3.10
  }
  if (grepl("SENTINEL", cname) == TRUE) { 
    yield <- 4.00  
  }
  if (grepl("PHOENIX", cname) == TRUE) { 
    yield <- 3.65 
  }
  if (grepl("NEW", cname) == TRUE) { 
    yield <- 3.55  
  }
  summary_df$Var <- cname
  summary_df$Rate <- yield #store the yield  
  master_df <- rbind(master_df, summary_df) #verticalstack the summary dataframe
}



```

```{r}
master_df %>% head(10) 
```

### Analyze Variance in Portfolio Compositions

```{r, fig.height=12, fig.width=12}
`%not in%` <- function (x, table) is.na(match(x, table, nomatch=NA_integer_))
manhattan_df <- subset(master_df,grepl("MANHATTAN", Var) == TRUE )
wula_df <- subset(master_df,grepl("WESTERN", Var) == TRUE )
others_df = subset(master_df, !(Var %not in% c("ATLANTIC COAST LIFE INSURANCE COMPANY","FIDELITY AND GUARANTY INSURANCE COMPANY", "LIBERTY BANKERS LIFE INSURANCE COMPANY", "SENTINEL SECURITY LIFE INSURANCE COMPANY" )))


theme_set(theme_dark())




g <- ggplot(others_df, aes(Sector, Weight, fill=Var)) + geom_bar(stat="identity", position="dodge", width=0.8, alpha = 0.5) + scale_fill_brewer(palette='Blues') + geom_bar(data=manhattan_df, stat="identity", position="dodge", width=0.2, alpha = 1, fill='tomato2', color='black') + theme(text = element_text(size = 11), axis.text.x = element_text(angle=90, hjust=1)) + ggtitle(" Manhattan vs. Top Competition (Manhattan=ORANGE)")

g
```

```{r, fig.height=12, fig.width=12}


g <- ggplot(others_df, aes(Sector, Weight, fill=Var)) + geom_bar(stat="identity", position="dodge", width=0.8, alpha = 0.5) + scale_fill_brewer(palette='Blues') + geom_bar(data=wula_df, stat="identity", position="dodge", width=0.2, alpha = 1, fill='tomato2', color='black') + theme(text = element_text(size = 11), axis.text.x = element_text(angle=90, hjust=1)) + ggtitle(" WULA vs. Top Competition (WULA=ORANGE)")
g
```


  
## 2. Granular Competitive Analysis


```{r}

p = list()
x = list() 


for (i in 1:length(psrt_data_list)) { 

  cname <- cnames[[i]]
  cdata  <- psrt_data_list[[i]]
  cdata <- processData(cdata, TRUE )
  p[[i]] <-  plot_treemap(cdata, cname)
  x[[i]] <- cdata
} 


```



```{r fig.width=10, fig.height=14}

do.call(grid.arrange, p)

```


```{r}
do.call(print, x)
```


```{r}
x[2]
write.csv(x[2], "Sentinel_Grouped_Holdings_July18.csv")
```